\documentclass{report}
\usepackage{color,soul,amsmath}
\providecommand{\abs}[1]{\lvert#1\rvert}
\providecommand{\norm}[1]{\lVert#1\rVert}

\title{Exploring the quantitative dependence of cell physiology, gene expression and proteome composition on growth rate in model microorganisms}
\author{Uri Barenholz}
\date{February 2013}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}
Single cell organisms are probably the simplest form of life accessible to science today.
However, complete understanding of their workings is still far from reached.
The ability of such organisms to self-replicate given a constant supply of simple substances is one of the simplest experimental setups that can be used to investigate the workings of a living system, albeit under a relatively simple environmental challenge.
The reasoning behind this experimental setting is that, at the absence of external changes to the environment, an organism should “converge” to a constant state (or replication cycle).
At this state the growth rate remains constant and the relative amounts of intracellular components (up to changes related to the cell cycle) are also maintained \cite{Campbell1957}.
Such setups have been used for decades \cite{Schaechter1958,Maaloe1969,Pedersen1978a} to explore the details of metabolism, resource allocation and composition of single celled organisms under so-called “balanced growth” or, more accurately, logarithmic balanced growth.
We note that the definition of logarithmic balanced growth is somewhat more restrictive than “(mid) exponential growth” as it implies not only constant growth rate but also semi-constant cell composition.

Advances in technology allowed for continuous increase in the resolution and breadth of measurements of the different components of cells under such conditions starting from the mere growth rate and wet/dry mass ratio, moving onto composition in terms of chemical elements, macromolecular composition and in recent years mRNA profiling and complete proteomic data using various methods.

As measurement techniques evolved and more details have been exposed, models and formulas of finer detail were devised to explain and predict various physiological aspects of growing cells \cite{Lerman2012,Karr2012}.
Advances in computational power allow for increasing complexity of simulation of various aspects of the cell and notably the introduction of Flux Balance Analysis and its various modifications as a tool for assessment of the metabolome in different states.
\section{Motivation for the understanding of physiology under balanced growth}
One of the basic features of any living system is its composition.
As such, attempts at identifying and characterizing the composition of living organisms have been a major part of biological research\cite{Schaechter1958,Bremer1987}.
Microorganisms have been playing a key role of this research ever since their discovery.
Balanced growth preserves cellular composition along time while increasing the amount of living matter.
As such, it is a very attractive target for research.
It is both relatively easy to study (due to its stability) and encompasses one of life's most interesting features - the ability to reproduce.

In the context of synthetic biology, one of the key promises of this field is to allow the synthesis of various high-value chemical compounds in an environmentally clean and economically efficient manner.
Realization of this vision is still in progress and the development of industrial processes using microorganisms at their core is a challenging task.
Lack of knowledge and tools that can assess the impact of the introduction of synthetic pathways into organisms is one of the obstacles being faced in this process.
The construction of a model for a functioning living organism (at balanced growth) is a valuable building block in the process of achieving such knowledge as it allows for assessment of effects of different modifications that can then be introduced into the system.
In recent years there have been few attempts to create such “whole cell” simulations \cite{Lerman2012,Karr2012,Reed2004,Schellenberger2011a}.
As these works suggest, enough data and understanding have been experimentally collected to allow valuable insights to be drawn out of such models.
\section{Current work}
In recent years there has been renewed intereset in the interplay between global factors and growth rate and gene expression.
In \cite{Klumpp2009a} it was shown that the protein concentration of a constitutively expressed gene is expected to decrease as a function of growth rate.
The model used measured dependence on growth rate of relevant cellular parameters: cellular copy number of the gene, transcription rate per gene copy, mRNA degradation rate, translation rate per mRNA, protein degradation rate and cell volume.
Despite the fact that the predicted number of protein molecules per cell increased, the increase in cell size due to growth rate outweighted this effect, reversing the result when protein concentration was considered.
The trend is much less clear when protein fraction out of the proteome is considered.
The model was compared, with relatively good agreement, with measurements of lacZ, TSase and OTCase reported from the mid-late 70's.
The rest of the article deals with the analysis of different regulatory mechanisms and their effects.
The authors conclude that ``Changes in gene expression need not reflect regulation''.
In \cite{Scott2010a} the impacts of modifying the translation rate (via anti-biotics) is concerned.
The authors derive formulas for describing the expected effects of modification of translation rate on growth rate and protein levels.
They also induce the production of unecessary proteins and monitor the effect on growth rate.
Their major results are:
\begin{enumeration}
\item Slower translation rate slows growth and reduces protein levels.
\item Production of unnecessary protein reduces growth rate in a proportional way.
\end{enumeration}
In \cite{Berthoumieux2013} an attempt is made to distinguish between the effects of global physiological state of the cell and specific regulation mechanisms.
The authors use an unregulated promoter (pRM from phage $\lambda$) as a benchmark and compare the activities of other promoters (fis, crp, acs) with it.
Data is obtained by monitoring the growth of a batch culture in 96-well plate and calculating the promoter activities throughout the growth phases (using placmids with fluorescent reporter genes).
For acs the authors conclude that specific regulation due to internal cAMP concentration has a major impact on promoter activity.
For fis and crp, the major impact on promoter activity is the global physiological state of the cell.
In \cite{loven2012} the issue of normalization of microarray measurements is discussed.
The authors claim that carful procedures need to be followed in order to correctly identify what part of the expression program changed vs. what measured changes are the result of changes in global cellular parameters.
This article demonstrates another aspect in which global physiological changes play a role in observed measurements and their analysis.
In \cite{Zaslaver2009a} the authors monitor the activity of a large set of promoters across different growth media and different growth phases.
They found that the fraction of ribosomal expression out of the entire set of expressed genes grows linnearly with growth rate.
Furthermore, they found that the overall distribution of promoter activities remains relatively constant but specific (metabolic) promoters present different activities depending on the growth media.
The authors present a model explaining the observed results for ribosomal promoters under a framework of optimal resource allocation.
According to the model, optimal allocation of resources between synthesizing ribosomal proteins and metabolic proteins in order to maximize growth rate, results in a ribosomal synthesis fraction that is linear in growth rate.
This result agrees well with the observed data.
In \cite{Levy2009} TODO

\section{Research aims}
\begin{enumerate}
\item Develop a model than can relate various physiological characteristics of the cell to the growth rate induced when at balanced growth.
The model will take into account various environmental conditions.
The model will focus on the following properties:
\begin{enumerate}
\item Overall macromolecular composition
\item Gene expression
\item Proteome composition
\item Cell size
\end{enumerate}
\item Collect measurements of promoter libraries and/or proteome composition under various growth conditions and use them to validate the model experimentally (to be done in collaboration with Leeat Yankielowicz-Keren from the Segal lab in S.Cerevisiae and Guy Aidelberg from the Alon lab for E.Coli).
\item Perform analysis on optimal resource utilization given both experimental expression measurements and insights from the model.
\end{enumerate}
\section{Workplan components}
\begin{enumerate}
\item Develop robust methods to calculate growth rate and gene expression level for batch cultures.
\item Collect accurate gene expression data under different growth conditions spanning a wide range of growth rates (in E.Coli and S. Cerevisiae , in collaboration with the Segal and Alon labs).
\item Investigate the connection between the expression of various subsystems/clusters of genes and the growth rate, possibly giving rationale for the correlations observed.
\item Integrate the data and correlations observed to a model that can be used to predict the physiological condition(macromolecular composition, gene expression levels) under new environmental conditions.
\item Compare the results to the predictions made by various existing models (specifically \cite{Lerman2012,Karr2012}).
\item Explore methods to combine thermodynamical data with existing whole-cell models.
\item *Investigate alternative experimental methods for measurement of growth rate and gene expression.
\end{enumerate}
\section{Preliminary results}
\subsection{Exploring balanced growth in batch cultures}
Although logarithmic balanced growth is well defined theoretically, it remains unclear if and how real cultures can exist in this phase.

One device that is commonly used for long term growth of cultures is a chemostat.
A comprehensive discussion of chemostats and their workings in beyond the scope of this document.
Briefly, chemostats replace a pre-defined fraction of a media in which a culture in grown by fresh media every some constant time interval.
They therefore allow continous growth over long periods of time.
However, it is unclear whether growth under such conditions is indeed balanced.
This is due to the fact that the culture converges to being under some limiting nutrient concentration.
When such a limiting concentration is reached, the growth of the culture is slowed down (prior to the next dilution step) and therefore the logarithmic balanced growth phase breaks.

Another way of apporximating balanced growth is by putting a small culture in a large volume of media, a situation that, for some time, allows the culture to grow with no limiting resource.
One of my aims in this research is to quantify the extent to which logarithmic balanced growth occures in such a setup.
The concrete experimental setup consists of a 96-well plate and a plate reader for measuring both the optical density and fluorescense of plasmids that are integrated into the bacteria.

Being able to identify logarithmic balanced growth given such a setup (if indeed it occures) is a valuable building block for extracting various quantitative data about different bacterial strains under different nutrients.
It should allow for good reproducibility and insightful comparisons of phenotipic behaviours that are hard to obtain otherwise.
\subsubsection{Growth rate assessment}
\label{growth-rate}
A key feature of logarithmic balanced growth is the maintenance of growth rate throughout this phase.
As such, accurate assessment of growth rate is required.

Despite being theoretically simple to calculate, in practice, obtaining accurate growth rate assessment over time requires overcoming various experimental obstacles.
These include, among others: accurate background determination and subtraction, detector noise, outlying measurements detection and removal and the detection of the period of time during which the growth rate is constant and maximal.

In our work we have developed a procedure that attempts to overcome these issues.
The key features of this procedure include background and detector noise determination and subtraction, use of high dilution ratio of the initial culture and use of Trimmed-Least-Squares robust fit methods \hl{cite reuseeaue} to overcome outlier measurements \hl{figure 1 - 4 panes - raw, background and threshold subtraction, trimming and log-scale, growth rate}.

We believe the combination of these techniques to give us a reliable, reproducable and versatile growth rate assessment method for a varaiety of growth media and organisms.

Where fluorescent reporter genes exist, we can estimate the accuracy of the measurement by comparing the results of executing this algorithm on the OD measurements with those obtained by executing it on fluorescense measurements.
If deviations between the two estimates occure we can conclude that either the growth is not balanced, or the procedure failed.

Further verification that the growth is indeed balanced during a phase can be done by verifying no correlation between the growth rate at a sliding subwindow throughout this time and the time itself.
\subsubsection{Protein level and promoter activity measurements}
Two features that are of interest for modeling the proteome of a cell, be it during balanced growth or otherwise, are the relative protein levels and the corresponding promoter activities.
Both features are hard, sometimes impossible, to measure directly without bias.

The fusion of fluorescent reporter genes to promoters in libraries is a common method to quantitatively deduce these features.
However, there is no canonical form or protocol for the inference of these features based on such experimental setup.
Two leading approaches involve estimating the amount of fluorescence per OD unit, which corresponds to protein abundance, and measuring the rate at which the fluorescence increases per OD unit, which corresponds to the production (or, more accurately, accumulation) rate of the protein.

While these two values are obtained from the same measurements, they are by no means equivalent.
If one assumes the protein level to follow the classic formula:
\[ \frac{dP}{dt}=\beta-\alpha P\]
(Where $\alpha$ is the dilution rate, equivalent to the growth rate, and $\beta$ is the production rate)
Then the steady-state concentration of the protein becomes:
\[ P_{st.st}=\frac{\beta}{\alpha}\]
Showing that the steady state concentration equales the production rate divided by the growth rate.

Monitoring these two properties throughout the growth phase of a culture gives us an opportunity to estimate how well the above stated mathematical relation holds and allows us to test if and how the expected convergence occures \hl{figure 2. 4 panels - fluorescence and OD growth rates on the same plot, activity over time, abundance over time and ratio of abundance to activity over time compared with growth rate}.
In order to achieve these goals we are developing procedures that allow us to extract accurate estimates for the fluorescence and OD values throughtout the growth phase.

Briefly, our procedure involves calculating linear fit on the background-subtracted logarithm of both the fluorescence and abosrbance measurements (in a similar manner to the one described in \ref{growth-rate}).
Comparison of the calculated slopes indicates how balanced the protein accumulation is compared with the growh rate.
The ratio between the calculated offsets of the fit lines measures the ratio of fluorescence to biomass which serves as a proxy for the protein abundance.
\subsection{A model for estimating the effect of resource allocation on protein accumulation rates}
When comparing promoter activity or gene expression data across different growth conditions one finds that nearly all promoters change their activity.
However - detailed examination of this change reveals that the genes can be subdivided into clusters and that the changes within each cluster (between two given conditions) can be encapsulated by a single scaling factor \cite{Leeat2013}.

In the following section we extend the analysis of the implications of having a limited production capacity, as is suggested in \cite{Leeat2013}.
We postulate that a large fraction of the changes observed in promoter activity and gene expression can be explained by a passive mechanism.
Putting this assumption to test and obtaining more data about protein production profiles under different conditions is one of the aims of our research.

As a start we describe the cell as a biomass-generating machine.
We make the assumption that, when at balanced growth, the ratio of protein to biomass remains constant (under the same growth condition.
This ratio is known to change somewhat under different growth conditions \cite{Bremer1987}.

We use the following notation:
\begin{description}
\item[$P$] Total protein amount (in culture, mass).
\item[$B$] Total biomass amount (of culture, mass).
\item [$dP/dt$] Rate of protein production (in culture, mass/time).
\item [$dB/dt$] Rate of biomass production (in culture, mass/time).
\item [$g$] Growth rate ($time^{-1}$).
\end{description}
We note that by definition:
\[ g=\frac{dB/dt}{B}\]
A corollary of the above equation is that for any biochemical entity $Q$ whose fraction out of the biomass $r_Q$ is constant (e.g. hl{XXX}), one gets that:
\subsubsection{Corollary}
If
\[\frac{Q}{B}=r_Q\]
is constant then:
\begin{equation}
\label{global-gr}
g=\frac{dB/dt}{B}=\frac{d(r_QB)/dt}{r_QB}=\frac{dQ/dt}{Q}
\end{equation}
Namely: The total rate of Q’s production divided by the total amount of Q is equal to the growth rate as long as Q’s ratio to the total biomass is constant.
This is, in fact, the definition of balanced growth \cite{Campbell1957}.

Applying equation \ref{global-gr} to the total protein production, denoting by $r$ the ratio of total protein to biomass, one gets that:
\[g=\frac{dP/dt}{P}=\frac{dP/dt}{rB}\rightarrow\frac{dP/dt}{B}=rg\]
The last term ($\frac{dP/dt}{B}$) has an interesting biological interpretation.
It is the protein production capacity of the culture per biomass unit.

We note that, again using equation \ref{global-gr}, if one assumes that the composition of the proteome remains constant under a given growth condition then for each protein $P_i$ it holds that:
\[g=\frac{dP_i/dt}{P_i}\]
That is, the production rate of protein $i$ divided by the amount of the protein in the biomass is equal to the growth rate.

This result suggests that the observed promoter activity of $P_i$ that is usually operatively defined as (though it is understood that this includes effects of the amount of the translation machinery and not only promoter effects):
\[A_i=\frac{dP_i/dt}{B}=\frac{dP_i/dt}{P_i}\frac{P_i}{B}=g\frac{P_i}{B}=rg\frac{P_i}{P}\]
Depends, by definition, both on the growth rate and on $P_i$’s fraction out of the total biomass (or out of the total protein in the biomass).
In light of this observation we aim to derive an “intrinsic”, growth-condition independent (meaning growth-rate and protein-to-biomass ratio independent), promoter activity measure.
A measure that, once determined, can be used to calculate the observed promoter activity for a given steady growth condition by taking into account the growth rate and proteome composition under this condition.
\subsubsection{The passive allocation model}
We explore a model where the majority of promoters are constitutively expressed, meaning they have no active regulation whatsoever but only some constant intrinsic “strength”.
We assume that regulated promoters can only have a few, discrete, transcriptionally controlled “strengths” (On-Off and rarely some intermediate states as well).
Despite its discreteness, such a model will exhibit continuous changes in the observed production rates due to the effects of growth-rate and protein composition described above.

According to our model the cell senses its growth environment and, for those promoters that can be regulated, decides in a discrete fashion on their expression strength (out of the limited set of strengths available for each such promoter).
The outcome of this process is a “task list” of genes to express, where each gene has an associated intrinsic strength.
For unregulated genes this strength is constant (per gene) and for regulated genes it is one of a finite subset of possible strengths.
Biologically, this list takes the form of active/ready-to-be-transcribed promoters.
This “task list” is then being processed by the transcription-translation machinery.
We assume this machinery operates at the maximum capacity, given the existing growth conditions, to express the genes.
The intrinsic strengths represent a partition function (or relative expression) of different genes with respect to one another and thus determines the final ratios between the different proteins.

Formally, we define the intrinsic activity level of promoter $i$ under condition $c$ as:
\[\Gamma^c_i=\delta_i+a^c_i\beta_i ; a^c_i\in \{a_i\}\]
where $\delta_i$ is the promoter basal level and, for each promoter with active regulation, there is a finite set of activation levels {0,...,1} out of which one,$a^c_i$, is used under condition $c$.
We note that such an intrinsic activity value, $\Gamma^c_i$, can be the combined result of RBS-sequence, RNA-polymerase affinity (that can be discretely modulated by the binding of transcription factors) or other (non-context dependent) mechanisms.

We postulate that the activity level of promoter $i$ under condition $c$ will be:
\begin{equation}
\label{cond-act}
A^c_i=r_cg_c\frac{\Gamma^c_i}{\norm{\vec{\delta}+\vec{a^c}I\vec{\beta}}}=r_cg_c\frac{\Gamma^c_i}{\sum_{j\in P}\Gamma^c_j}
\end{equation}
(Where the middle term uses vector notation with $I$ being the identity matrix).

An important point one needs to pay attention to here is that the composition of the proteome under condition c (namely the denominators of the middle and right hand sides) may change between different growth conditions.
However, these changes are entirely encapsulated in the condition specific $\vec{a^c}$ vector of the discrete activation levels.
One of our aims in this research is to try and test to what extent this assumption holds.
\subsubsection{Deriving an explicit expected activity level}
In order to test the proposed model we need to calculate what the expected activity level under our model should be.
The term we derived above includes the sum over the static activity levels of all promoters under condition c.
Using this equation is therefore somewhat impractical as the sum needs to be determined.
We continue to show how this term can be eliminated to get an observed activity level that depends only on $\Gamma^c_i$ itself.

We make two assumptions:
\begin{enumerate}
\item The translation rate is constant.
\item There is a set of genes, $G_t$ , that produce the translation machinery, and these genes are unregulated (meaning their expression level is not under active regulation).
\end{enumerate}
Under these assumptions we denote the sum of intrinsic strengths of the unregulated, translation-machinery producing genes by:
\[\Omega=\sum_{i \in G_t}\Gamma_i=\sum_{i \in G_t}\delta_i\]
Note that, as these genes are unregulated, no specific activity levels get involved in the sum.
This term is therefore constant across different conditions.

Next we make the following observation:
As these genes encode the translation mechanism proteins then, assuming constant translation rate, there is a constant time $t_{tr}$ that this mechanism takes to translate its own genes.
For simplicity one can think about the time it takes a ribosome to translate the equivalent of its own proteins.
Note that when a cell doubles itself, every ribosome needs to translate at least the equivalent of its own proteins plus some portion of the rest of the proteome.
Therefore, the total time it takes the translation mechanism to produce a desired proteome under condition c, and thus to double the biomass, assuming constant translation rate, is:
\[\tau_c=\frac{\sum_{j\in P}\Gamma^c_j}{\Omega}t_{tr}\]
Remebering that by  definition $g=\ln(2)/\tau$ yields:
\[g_c=\ln(2)\frac{\Omega}{t_{tr}\sum_{j\in P}\Gamma^c_i}\]
Substituting this result in the activity level equation (equation \ref{cond-act}) for condition c we get that:
\begin{equation}
\label{ind-act}
A^c_i=r_c g_c \frac{\Gamma^c_i}{\sum_{j\in P}\Gamma^c_j}=r_c g_c \frac{\Gamma^c_i}{\Omega}\frac{\Omega}{\sum_{j\in P}\Gamma^c_j}=r_c g_c^2\frac{t_{tr}}{\ln(2)}\frac{\Gamma^c_i}{\Omega}
\end{equation}
Leaving us with an observed activity level that depends linearly on the protein/biomass ratio, the square of the growth rate, some condition-independent constants and the condition specific intrinsic strength of the relevant promoter.
\subsubsection{Accounting for variable translation rate}
As noted above, the preceding analysis assumes the translation rate remains constant under different growth rates.
We note that this assumption was used in exactly one place, that is, in the assertion that $t_{tr}$ is constant, as that is the only place where there is a direct connection between intrinsic strengths and translation times.
Thus, relieving this assumption and instead assuming that the translation rate is some function of the growth rate, $f$, one gets that the time the translation mechanism takes to translate itself is simply:
\begin{equation}
t_{tr}=f(g_c)
\end{equation}
Incorporating this term into the activity level equation (equation \ref{ind-act}) we conclude that in such cases:
\begin{equation}
A^c_i=r_c g_c \frac{\Gamma^c_i}{\sum_{j\in P}\Gamma^c_j}=r_c g_c \frac{\Gamma^c_i}{\Omega}\frac{\Omega}{\sum_{j\in P}\Gamma^c_j}=r_c g_c^2\frac{f(g_c)}{\ln(2)}\frac{\Gamma^c_i}{\Omega}
\end{equation}
It is reasonable to expect the dependence of the translation rate on the growth rate to be a monotonic, slowly decreasing function.
In such a case, the resulting final relation of the production rate to the growth rate should therefore be a sub-quadratic relation.
\subsubsection{Validation of the model for Ribosomal proteins}
As a test case for the model and our ability to set up an experimental system sensitive enough to test it, Leeat Y. assembled a plate spanning various media with a strain of S. Cerevisiae expressing GFP fused to a ribosomal promoter.
Our analysis shows that, indeed, the accumulation rate scales like the growth rate squared.
One should note that there are other explanations for this dependence and therefore other promoters will be tested as well.
\subsection{Correlating growth rate and the protein to biomass ratio}
A simple calculation shows that the protein to biomass ratio is expected to change as a function of the growth rate.
As Ribosomes are composed of 1:2 ratio of protein to RNA then, under the assumption of constant translation rate, one expects that the ribosomal protein fraction out of the proteome will be:
\[\alpha_R=\frac{t_{tr}}{\tau}=t_{tr}g/\ln(2)\]
And therefore, the ribosomal RNA (rRNA) amount relative to the protein amount should follow:
\[\alpha_{rRNA}=2\alpha_R=2t_{tr}g/\ln(2)\]
Meaning it is an increasing function of $g$.
Experimental data supports this model as it shows increase in RNA/protein ratio as growth rate increases.
Note that the other factors needed for calculation of the global macromolecular composition are the relative fractions of other cellular components (lipids, DNA, small metabolites, all of which are expected to stay relatively constant across different growth rates) and the relative fraction of rRNA out of the total RNA in the cell (which is commonly cited as being ~ 85\% as long as growth rates are not extremely slow \cite{Bremer1987,goldberger1979biological}).

We aim to measure the RNA to protein ratio to check the validity of this prediction.
\bibliographystyle{plain}
\bibliography{library}
\end{document}
